## Queries and Statements

```{r, include=FALSE}
con %>% dbDisconnect()
con <- issue_con("demo")
```

This section uses the following tables as an example.
```{r queries-section-init, collapse=TRUE}
con %>% dbWriteTable("mtcars", mtcars, overwirte = TRUE)
con %>%
    dbReadTable("mtcars") %>%
    head()
```

### `dbGetQuery()`

- Executes a `SELECT` query written as a string.
- Returns the result, which is not stored as a result set contrary to `dbSendQuery()`.
- Comes with a default implementation that calls `dbSendQuery()` -> `dbFetch()` -> `dbClearResult()`.
- Returns a dataframe even if no rows are returned.
- Additional arguments
	- `n = -1`: specifies the number of rows to be fetched. `n = Inf` gives you the full result, `n = 0` the fully typed columns.
	- `param`: allows passing query parameters, see `dbBind()` for details.
	- `immediate`: see official reference.

You don't have to add `;` at the end of the query.
```{r getquery1, collapse=TRUE}
con %>% dbGetQuery("SELECT * FROM mtcars WHERE cyl >= 5", n = 5)
```

Sclar values in a query can be parameterized.

```{r getquery-param, collapse=TRUE}
con %>% dbGetQuery(
    "SELECT * FROM mtcars WHERE cyl = $1",
    params = list(1:5),
    n = 5
)
```

`dbGetQuery()` rejects multiple queries.

```{r cannot-get-multiple-queries, error=TRUE, collapse=TRUE}
con %>% dbGetQuery("SELECT * FROM mtcars WHERE cyl >= 5; SELECT 3.14 AS pi")
```

### `dbExecute()`

- Executes a statement. The result is not stored as a result set.
- Returns the number of rows affected. 
- Comes with a default implementation that calls `dbSendStatement()` -> `dbGetRowsAffected()` -> `dbClearResult()`.
- Additional arguments `params` and `immediate` are available as in `dbGetQuery()`.

```{r dbExecute, collapse=TRUE}
# newly create cars table
con %>% DBI::dbWriteTable("cars", cars %>% head(1), overwrite = TRUE)
con %>% dbReadTable("cars") # only 1 row

con %>% dbExecute(
    "INSERT INTO cars (speed, dist) VALUES ($1, $2)",
    params = list(1:4, 6:9)
)
con %>% dbReadTable("cars") # rows added
```

`dbExecute()` rejects multiple statements.

```{r cannot-executes-multiple-statements, error=TRUE, collapse=TRUE}
con %>% dbExecute("INSERT INTO cars (speed, dist) VALUES (7,21); DROP TABLE cars")
```

### `dbSendQuery()`

- Submits and synchronously executes the SQL query. Should be followed by `dbFetch()` and `dbClearResult()` to get and clear the pooled result. See [The data retrieval flow section in official reference](https://dbi.r-dbi.org/reference/dbsendquery).
- Returns an S4 object that inherits from DBIResult, called a result set.
- The returned result object should be stored in a variable.
- Additional arguments `params` and `immediate`.

Exactly the same parameterization method works as in `dbGetQuery()`.

```{r send-query-example1, collapse=TRUE}
rs <- con %>% dbSendQuery(
    "SELECT * FROM mtcars WHERE cyl = $1",
    params = list(4L)
)
rs %>% dbFetch(n = 5)
rs %>% dbClearResult()
```

`dbSendQuery()` gives us an alternative way of writing a parameterized query.

```{r send-query-example2, collapse=TRUE}
rs <- con %>% dbSendQuery(
    "SELECT * FROM mtcars WHERE cyl = $1"
)
rs %>% dbBind(list(4L))
rs %>% dbFetch(n = 5)
rs %>% dbClearResult()
```

### `dbSendStatement()`

- Submits and synchronously executes the SQL data manipulation statement (e.g., UPDATE, DELETE, INSERT INTO, DROP TABLE, ...).
- Returns a result object as in `dbSendQuery()`.
- The returned result set should be stored in a variable.
- A part of `dbExecute()`.
- Additional arguments `params` and `immediate`.

A typical parameterization through `params = list(...)` works.

```{r dbSendStatement-param, collapse=TRUE}
# overwirte cars table
con %>% DBI::dbWriteTable("cars", cars %>% head(1), overwrite = TRUE)
con %>% dbReadTable("cars")

# insert statement with typical parameterization
rs <- con %>% dbSendStatement(
    "INSERT INTO cars (speed, dist) VALUES ($1, $2)",
    params = list(1:4, 5:8)
)
dbClearResult(rs)
con %>% dbReadTable("cars")
```

You can also pass parameters by using `dbBind()`.

```{r dbSendStatement-param2, collapse=TRUE}
# overwirte cars table
con %>% DBI::dbWriteTable("cars", cars %>% head(1), overwrite = TRUE)
con %>% dbReadTable("cars")

# pass multiple parameter using dbBind()
rs <- dbSendStatement(
    con,
    "INSERT INTO cars (speed, dist) VALUES ($1, $2)"
)
dbBind(rs, list(2:3, 4:5))
dbBind(rs, list(6L, 7L))
dbClearResult(rs)
dbReadTable(con, "cars")
```

### `dbBind()`

- Binds placeholders in a parameterized/prepared statement to actual values.
- Returns a result set invisibly.
- Is intented to be called before calling `dbFetch()` or `dbGetRowsAffected()`.
- The placeholder syntax depends on DBMS.
	- `?`: RMySQL and RSQLite
	- `$1`: RPostgres and RSQLite
	- `:name` and `$name`: RSQLite.

```{r bind-params, collapse=TRUE}
con %>% dbWriteTable("iris", iris %>% head(), overwrite = TRUE)
rs <- con %>% dbSendQuery("SELECT * FROM iris WHERE 'Petal.Width' > $1")
rs %>% dbBind(list(3.0))
dbFetch(rs)
```

See `dbSendQuery` and `dbSendStatement` for more examples.

### `dbFetch(), fetch()`

- Fetches the next `n` elements (rows) from the result set.
- Returns a dataframe regardless of the result.
- For all new code, use `dbFetch()` not `fetch()` which is provided just for compatibility with older DBI-client. 

`dbFetch()` fetches unfetched results, as the next example shows.

```{r dbFetch-example1, collapse=TRUE}
rs <- con %>% dbSendQuery("SELECT mpg, cyl FROM mtcars LIMIT 10")
while (rs %>% dbHasCompleted() %>% isFALSE()) {
    chunk <- dbFetch(rs, 2)
    print(chunk)
}

dbClearResult(rs)
```

Not all result objects can be fetched.
One such example is the result of query with a unbounded placeholder.

```{r dbFetch-counter-example1, error=TRUE ,collapse=TRUE}
rs <- con %>% dbSendQuery("SELECT mpg, cyl, disp FROM mtcars WHERE cyl = $1")
rs %>% dbFetch(n = 5) # placeholder should be bound before being fetched
rs %>% dbBind(list(6L))
rs %>% dbClearResult()
```

More trivial result is a cleared result.

```{r dbFetch-counter-example2, error=TRUE ,collapse=TRUE}
rs %>% dbClearResult()
rs %>% dbFetch()
```

### `dbGetRowCount()`

- Counts the total number of rows **actually returned** by `dbFetch()` for the input result object.
- Returns a scalar number.

```{r get-row-count-example, collapse=TRUE}
rs <- dbSendQuery(con, "SELECT * FROM mtcars")

rs %>% dbGetRowCount() # initial value of the method

count_fetched_rows <- 0
while (!dbHasCompleted(rs)) {
    count_fetched_rows <- count_fetched_rows + nrow(rs %>% dbFetch(10))
}

count_fetched_rows # total number of rows fetched
rs %>% dbGetRowCount()

# fetching empty row has no effect on the value of dbGetRowCount()
rs %>% dbFetch()
rs %>% dbGetRowCount()

rs %>% dbClearResult()
```

`dbGetRowCount()` cannot count a cleared result.

```{r get-row-count-example2, error=TRUE,collapse=TRUE}
rs %>% dbGetRowCount()
```

### `dbGetRowsAffected()`

- Gets the number of rows affected by a data manipulation statement issued with dbSendStatement()
- Returns a scalar number, integer or numeric.

```{r dbGetRowsAffected-example1, collapse=TRUE}
con %>% dbWriteTable("cars", cars, overwrite = TRUE)
rs <- con %>% dbSendStatement("DELETE FROM cars WHERE dist >= 20")
rs %>% dbGetRowsAffected()
rs %>% dbClearResult()
```

The following example shows that `dbGetRowsAffected()` increases when placeholders get bounded, not when placeholders are placed.

```{r dbGetRowsAffected-example2, collapse=TRUE}
con %>% dbWriteTable("cars", cars, overwrite = TRUE)
rs <- con %>% dbSendStatement("DELETE FROM cars WHERE speed = $1 AND dist = $2")
rs %>% dbGetRowsAffected()
rs <- rs %>% dbBind(list(7, 21))
rs %>% dbGetRowsAffected() # there is no rows with that value
rs %>% dbClearResult()
```

### `dbGetStatement()`

- Gets the statement that was passed to `dbSendQuery()` or `dbSendStatement()`. It remains unchanged after placeholders are bounded afterwards.
- Returns a string statement.

```{r get-statement, collapse=TRUE}
con %>% dbWriteTable("cars", cars, overwrite = TRUE)
rs <- con %>% dbSendStatement("SELECT * FROM cars WHERE dist >= $1")
stm <- rs %>% dbGetStatement()
stm %>% class()
stm %>% print()
rs %>% dbBind(list(25))
rs %>% dbGetStatement() # unchanged!
rs %>% dbFetch(n = 5)
rs %>% dbClearResult()

# reuse the derived statement
rs <- con %>% dbSendStatement(stm)
rs %>% dbBind(list(35))
rs %>% dbFetch(n = 5)
rs %>% dbClearResult()
```

### `dbHasCompleted()`

- Tests if the associated operation has completed. 
- Returns a logical. `TRUE` if there is nothing more to be fetched and `FALSE` if otherwise.

`SELECT` statement completes when all results, if any, have been fetched.

```{r dbHasCompleted-example1, collapse=TRUE}
con %>% dbWriteTable("mtcars", mtcars, overwrite = TRUE)
rs <- con %>% dbSendQuery("SELECT * FROM mtcars")
rs %>% dbHasCompleted()
rs %>%
    dbFetch(n = 20) %>%
    invisible() # discard
rs %>% dbHasCompleted()
rs %>%
    dbFetch() %>%
    invisible() # discard
rs %>% dbHasCompleted()
rs %>% dbClearResult()
```

`SELECT` statement with no result rows completes as soon as the query is executed. Consequently, there is no need to fetch it to get it completed.

```{r dbHasCompleted-example2, collapse=TRUE}
con %>% dbWriteTable("mtcars", mtcars, overwrite = TRUE)
rs <- con %>% dbSendQuery("SELECT * FROM mtcars WHERE cyl = 5")
rs %>% dbHasCompleted()
rs %>% dbClearResult()
```

`DELETE`, `INSERT` and `UPDATE` statements get completed immediately after execution because there is originally nothing to fetch.

```{r dbHasCompleted-example3, collapse=TRUE}
con %>% dbWriteTable("cars", cars, overwrite = TRUE)
rs <- con %>% dbSendStatement("DELETE FROM cars WHERE speed >= 8")
rs %>% dbHasCompleted()
rs %>% dbClearResult()
```

### `dbColumnInfo()`

- Returns `data.frame` that describes the output of a query.
- A returned `data.frame` at least includes `"name"` and `"type"` followed by optional columns that starts with a dot such as `".oid"` and `".typname"` depending on DBMS.

```{r column-info-example, collapse=TRUE}
rs <- con %>% dbSendQuery(
    "
	SELECT 3.14 AS pi,
	0 AS unit_addition,
	'shena' AS author,
	current_date AS today
	"
)
info <- rs %>% dbColumnInfo()
info %>% class()
info %>% mode()
info %>% print()
rs %>% dbFetch(n = 0)
rs %>% dbClearResult()
```

### `dbClearResult()`

- Frees all resources (local and remote) associated with a result set.
- Returns `TRUE` invisibly.
- MUST be called `dbSendQuery()` and `dbSendStatement()`.

Rposgres automatically cleans up the existing result setes associated connnection when a new `dbSendQuery()` and `dbSendStatement()` is called. So, Rposgres user don't need to pay much attention to this method.

```{r clear-result-example, collapse=TRUE, warning=TRUE}
rs <- con %>% dbSendQuery("SELECT 3.14 AS pi")
rs %>% dbFetch()
rs %>% dbClearResult()
rs %>% dbClearResult() # clearing a cleared result again causes warning
```

If you are using RPosgres, automatic clearing works.

```{r clear-result-example2, collapse=TRUE, warning=TRUE}
rs <- con %>% dbSendQuery("SELECT 3.14 AS pi")
rs <- con %>% dbSendQuery("SELECT 2.71 AS napier") # this call induces auto-cleaning
rs %>% dbFetch()
rs %>% dbClearResult()
```

The automatic cleaning does not work if you are working on different connnection variables. Such situation should be avoided or be handled with care in general.

```{r clear-result-example3, collapse=TRUE, warning=TRUE}
con1 <- issue_con("demo") # new connection
rs <- con %>% dbSendQuery("SELECT 3.14 AS pi")
rs1 <- con1 %>% dbSendQuery("SELECT 2.71 AS napier") # query with the new connection
rs %>% dbFetch() # no automatic cleaning
rs1 %>% dbFetch()

rs1 %>% dbClearResult()
rs %>% dbClearResult()
con1 %>% dbDisconnect()
```