## Queries and Statements

```{r, include=FALSE}
con <- refresh_con(con)
```

This section uses the following tables as an example.
```{r }
con %>% dbWriteTable("mtcars", mtcars, overwirte = TRUE)
con %>% dbListTables()
```

### `dbGetQuery()`

- Executes a `SELECT` query written as a string.
- Returns the result, which is not stored as a result set contrary to `dbSendQuery()`.
- Comes with a default implementation that calls `dbSendQuery()` -> `dbFetch()` -> `dbClearResult()`.
- Returns a dataframe even if no rows are returned.
- Additional arguments
	- `n = -1`: specifies the number of rows to be fetched. `n = Inf` gives you the full result, `n = 0` the fully typed columns.
	- `param`: allows passing query parameters, see `dbBind()` for details.
	- `immediate`: see official reference.


You don't have to add `;` at the end of the query.
```{r getquery1}
con %>% dbGetQuery("SELECT * FROM mtcars WHERE cyl >= 5", n = 5)
```

```{r getquery-param}
con %>% dbGetQuery(
    "SELECT * FROM mtcars WHERE cyl = $1",
    params = list(1:5),
    n = 5
)
```

### `dbExecute()`

- Executes a statement. The result is not stored as a result set.
- Returns the number of rows affected. 
- Comes with a default implementation that calls `dbSendStatement()` -> `dbGetRowsAffected()` -> `dbClearResult()`.
- Additional arguments `params` and `immediate` are available as in `dbGetQuery()`.

```{r dbExecute, collapse=TRUE}
# newly create cars table
con %>% DBI::dbWriteTable("cars", cars %>% head(1), overwrite = TRUE)
con %>% dbReadTable("cars") # only 1 row

con %>% dbExecute(
    "INSERT INTO cars (speed, dist) VALUES ($1, $2)",
    params = list(1:4, 6:9)
)
con %>% dbReadTable("cars") # rows added
```

### `dbSendQuery()`

- Submits and synchronously executes the SQL query. Should be followed by `dbFetch()` and `dbClearResult()` to get and clear the pooled result. See [The data retrieval flow section in official reference](https://dbi.r-dbi.org/reference/dbsendquery).
- Returns an S4 object that inherits from DBIResult, called a result set.
- The returned result object should be stored in a variable.
- Additional arguments `params` and `immediate`.

Exactly the same parameterization method works as in `dbGetQuery()`.

```{r send-query-example1, collapse=TRUE}
rs <- con %>% dbSendQuery(
    "SELECT * FROM mtcars WHERE cyl = $1",
    params = list(4L)
)
rs %>%
    dbFetch() %>%
    head()
rs %>% dbClearResult()
```

`dbSendQuery()` gives us an alternative way of writing a parameterized query.

```{r send-query-example2, collapse=TRUE}
rs <- con %>% dbSendQuery(
    "SELECT * FROM mtcars WHERE cyl = $1"
)
rs %>% dbBind(list(4L))
rs %>%
    dbFetch() %>%
    head()
rs %>% dbClearResult()
```

### `dbSendStatement()`

- Submits and synchronously executes the SQL data manipulation statement (e.g., UPDATE, DELETE, INSERT INTO, DROP TABLE, ...).
- Returns a result object as in `dbSendQuery()`.
- The returned result set should be stored in a variable.
- A part of `dbExecute()`.
- Additional arguments `params` and `immediate`.

A typical parameterization through `params = list(...)` works.

```{r dbSendStatement-param, collapse=TRUE}
# overwirte cars table
con %>% DBI::dbWriteTable("cars", cars %>% head(1), overwrite = TRUE)
con %>% dbReadTable("cars")

# insert statement with typical parameterization
rs <- con %>% dbSendStatement(
    "INSERT INTO cars (speed, dist) VALUES ($1, $2)",
    params = list(1:4, 5:8)
)
dbClearResult(rs)
con %>% dbReadTable("cars")
```

You can also pass parameters by using `dbBind()`.

```{r dbSendStatement-param2, collapse=TRUE}
# overwirte cars table
con %>% DBI::dbWriteTable("cars", cars %>% head(1), overwrite = TRUE)
con %>% dbReadTable("cars")

# pass multiple parameter using dbBind()
rs <- dbSendStatement(
    con,
    "INSERT INTO cars (speed, dist) VALUES ($1, $2)"
)
dbBind(rs, list(2:3, 4:5))
dbBind(rs, list(6L, 7L))
dbClearResult(rs)
dbReadTable(con, "cars")
```

### `dbBind()`

- Binds placeholders in a parameterized/prepared statement to actual values.
- Returns a result set invisibly.
- Is intented to be called before calling `dbFetch()` or `dbGetRowsAffected()`.
- The placeholder syntax depends on DBMS.
	- `?`: RMySQL and RSQLite
	- `$1`: RPostgres and RSQLite
	- `:name` and `$name`: RSQLite.

```{r bind-params, collapse=TRUE}
con %>% dbWriteTable("iris", iris %>% head(), overwrite = TRUE)
rs <- con %>% dbSendQuery("SELECT * FROM iris WHERE 'Petal.Width' > $1")
rs %>% dbBind(list(3.0))
dbFetch(rs)
```

See `dbSendQuery` and `dbSendStatement` for more examples.

### `dbFetch(), fetch()`

- Fetches the next `n` elements (rows) from the result set.
- Returns a dataframe regardless of the result.
- For all new code, use `dbFetch()` not `fetch()` which is provided just for compatibility with older DBI-client. 

`dbFetch()` fetches unfetched results, as the next example shows.

```{r dbFetch-example1, collapse=TRUE}
rs <- con %>% dbSendQuery("SELECT mpg, cyl FROM mtcars LIMIT 10")
while (rs %>% dbHasCompleted() %>% isFALSE()) {
    chunk <- dbFetch(rs, 2)
    print(chunk)
}

dbClearResult(rs)
```

Not all result objects can be fetched.
One such example is the result of query with a unbounded placeholder.

```{r dbFetch-counter-example1, error=TRUE ,collapse=TRUE}
rs <- con %>% dbSendQuery("SELECT mpg, cyl, disp FROM mtcars WHERE cyl = $1")
rs %>% dbFetch(n = 5) # placeholder should be bound before being fetched
rs %>% dbBind(list(6L))
rs %>% dbClearResult()
```

A more trivial result is a cleared result.

```{r dbFetch-counter-example2, error=TRUE ,collapse=TRUE}
rs %>% dbClearResult()
rs %>% dbFetch()
```

### `dbGetRowCount()`

- Counts the total number of rows **actually returned** by `dbFetch()` for the input result object.
- Returns a scalar number.

```{r get-row-count-example, collapse=TRUE}
rs <- dbSendQuery(con, "SELECT * FROM mtcars")

rs %>% dbGetRowCount() # initial value of the method

count_fetched_rows <- 0
while (!dbHasCompleted(rs)) {
    count_fetched_rows <- count_fetched_rows + nrow(rs %>% dbFetch(10))
}

count_fetched_rows # total number of rows fetched
rs %>% dbGetRowCount()

# fetching empty row has no effect on the value of dbGetRowCount()
rs %>% dbFetch()
rs %>% dbGetRowCount()

rs %>% dbClearResult()
```

`dbGetRowCount()` cannot count a cleared result.

```{r get-row-count-example2, error=TRUE,collapse=TRUE}
rs %>% dbGetRowCount()
```

### `dbGetRowsAffected()`

The number of rows affected

### `dbGetStatement()`

Get the statement associated with a result set

### `dbHasCompleted()`

Completion status

### `dbColumnInfo()`

Information about result types

### `dbClearResult()`

Clear a result set