## SQL

### `SQL()`

- Hndles SQL quoting.
- Returns a `SQL` class object. 
- See `dbQuoteIdentifier()` or `dbQuoteLiteral()` for similar (but different) quoting methods.
- See `sqlInterpolate()` for an application.
- `glue_sql()` in glue packages is also useful to deal with SQL escaping needs.

```{r sql-method-exampl1, collapse=TRUE}
con <- refresh_con(con)
str <- c("SELECT", "WHERE")
s <- SQL(str)
s %>% class()
s %>% print()

SQL(s) %>% print() # unchanged

dbQuoteLiteral(con, s) %>% print() # SQL vectors pass through dbQuoteLiteral()

lt <- dbQuoteLiteral(con, str)
lt %>% print()

dbQuoteLiteral(con, lt) %>% print() # double escaping doesn't happen
```

### `dbDataType()`

- Describes the corresponding SQL data type for the DBMS.
- Returns a character string.
- Works without valid connection.

```{r sql-data-type, error=TRUE, collapse=TRUE}
con %>% dbDataType(Sys.Date()) # specify DB by connection
# works with a disconnected connection variable
con %>% dbDisconnect()
con %>% dbDataType(Sys.time())
con %>% dbDataType(Sys.time() - as.POSIXct(Sys.Date()))
con %>% dbDataType(c("pi", "e"))
con %>% dbDataType(list(raw(10), raw(20)))
con %>% dbDataType(I(1))
con %>% dbDataType(iris)
con %>% dbDataType(NA)
con %>% dbDataType(NULL) # error
```

You can also pass a `DBIDriver` object to specify the target DB.

```{r sql-data-type2,collapse=TRUE}
dbDataType(RPostgres::Postgres(), 1) # specify DB by driver
dbDataType(RPostgres::Postgres(), iris) %>% class()
```

### `dbQuoteIdentifier()`

### `dbQuoteLiteral()`

### `sqlInterpolate()`

- Create a SQL statement from a query string with a greater flexibility than `dbExecute()` or `dbGetQuery()`.
- Returns SQL of `SQL` class.
- Accepts placeholders with one of two syntaxes. They are bounded by `.dots = list()`. 
	- `?,...,?`: bounded by an unnamed list.
	-  `?name1,...,?name2`: bounded by a named list.
- All values in `.dots = list()` will be first escaped with `dbQuoteLiteral()` to protect against SQL injection.

[DBI references](https://dbi.r-dbi.org/reference/sqlinterpolate) has many examples including several anti-patterns that should be taken care of.

```{r sql-interpolate-example1, collapse=TRUE}
con <- issue_con("demo")
sql <- "SELECT * FROM ?"
con %>% sqlInterpolate(
    sql = sql,
    .dots = list("mtcars" %>% dbQuoteIdentifier(con, .))
)
```

```{r sql-interpolate-example2, collapse=TRUE}
sql <- "SELECT ?p AS ?name"
stm <- con %>%
    sqlInterpolate(
        sql = sql,
        p = 3.14,
        name = "pi" %>% dbQuoteIdentifier(con, .)
    )
stm %>% print()

con %>% dbGetQuery(stm)
```

```{r sql-interpolate-example3, collapse=TRUE}
sql <- "DISCARD ?var"
all <- FALSE

con %>% sqlInterpolate(
    sql,
    var = if_else(all, "ALL", "TEMPORARY") %>% SQL()
)
```