---
title: "Code Examples for DBI and RPostgres"
author: "Shena"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
github-repo: 'Shena4746/datascience-100knocks-preprocess-R'
---

```{r setup, eval = TRUE, include=FALSE}
rm(list = ls())
gc(reset = TRUE)
gc(reset = TRUE)
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
```

# What's this? 

From [README.md in Github repository]():

This repository is the R markdown source for the article [Code Examples for DBI and RPostgres](https://shena4746.github.io/code-examples-dbi-rpostgres/).
Assuming you are using DBI with Postgres, it provides a brief overview of each DBI function and various code examples that illustrates how it is typically used and how it behaves.

It is also intended to document minor tips that are not as well-documented as DBI official references. One important example is the actual behaviour of DBI functions.
As implied in [DBI specification](https://dbi.r-dbi.org/articles/spec), the behaviour of DBI functions, which depends on the R/DBMS implementation, may sometimes be slightly different from what is expected from the DBI documentation.
In such cases, the user is required to learn where the gaps exist and how they are different, typically through ad-hoc experimentation. This is exactly where this article grows out of. As a result, the content tends to focus on tricky aspects at times. If you would like a more balanced explanation, then the official DBI references should still be your first choice.

# About DBI and RPostgres

[DBI specification](https://dbi.r-dbi.org/articles/spec) has a rich explanation. Here is a brief summary of what DBI and RPostgres are and their relationship: 

- DBI (shor fo the database inteface) defines virtual classes such as DBIDriver, DBIConnection, and DBIResult, which is intended to serve as an interface for communitacation between R and R/DBMS.
- The connection to indivisual DBMS is provides by so-called DBI backend packages such as RSQLite, RPostgres, RMySQL, RSQLServe, etc. It also provides implementation for the DBI base classes mentioned above.

What is sometimes confusing is the fact that the functionlaity of the base class might be extended or/and might not be fully supported by a particular backend. Also, the backed class is not so well documented as DBI. Unfortunately, these things also apply to RPostgres, which is one of the reason why some "experimentation" are needed as mentioned earlier.

## List of Functions

```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library("RPostgres")
library("DBI")
library("dbplyr")
library("dplyr")
```

The following contents rely on the following packages.
```{r versions-of-libraries, collapse=TRUE}
R.version.string
packageVersion("RPostgres")
packageVersion("DBI")
packageVersion("dbplyr")
packageVersion("dplyr")
packageVersion("purrr")
```

### Functions in DBI 

```{r DBI-functions, collapse=TRUE}
# functions in dbi package
dbi <- ls("package:DBI")
dbi %>% length() # how many ?
dbi %>% print()
```

### Functions in RPostgres

```{r RPostgres-functions, collapse=TRUE}
# functions in RPostgres package
rpos <- ls("package:RPostgres")
rpos %>% length() # how many ?
rpos %>% print()
```

### Functions with duplicated names

```{r DBI-RPostgres-common-functions, collapse=TRUE}
# common functions in DBI and RPostgres
common <- base::intersect(dbi, rpos)
common %>% length() # how many ?
common %>% print()
```

### Funcions only in DBI

```{r DBI-unique-functions, collapse=TRUE}
# functions only in DBI
only_in_dbi <- base::setdiff(dbi, rpos)
only_in_dbi %>% length() # how many ?
only_in_dbi %>% print()
```

### Funcions only in RPostgres

```{r RPostgres-unique-functions, collapse=TRUE}
# functions only in RPostgres
only_in_rpos <- base::setdiff(rpos, dbi)
only_in_rpos %>% length() # how many ?
only_in_rpos %>% print()
```

## Basic Usage

Before we look at each function, let us describe a typical work flow with these packages.
You may sefaly skip this section as it is independent of what follows.

This section assumes that you have access to a postgres database. If necessary, run the following command to create a new database named 'demo'.

```{bash start-postgres, eval=FALSE}
sudo service postgresql start
```

```{bash, eval=FALSE}
createdb -p 5432 --owner=your-owner-name --host=localhost --username=your-username demo
```

The following code defines the `con` variable which holds a establised connection to your database. 

```{r example-connect, eval = FALSE}
con <- DBI::dbConnect(
    drv = RPostgres::Postgres(),
    host = "localhost",
    port = 5432,
    dbname = "your-database-name",
    user = "your-user",
    password = "your-password"
)
```

You should not write this way outside this demo. See \@ref(dbConnect).

```{r example-issue-con, include=FALSE}
dw <- config::get("demo")
con <- DBI::dbConnect(
    drv = RPostgres::Postgres(),
    host = dw$host,
    port = dw$port,
    dbname = dw$dbname,
    user = dw$user,
    password = dw$password
)
```

```{r remove-all-tables, include=FALSE}
con %>%
    dbListTables() %>%
    purrr::map_lgl(~ dbRemoveTable(con, .x)) %>%
    all()
```


Let us observe the list of tables in the connected database.

```{r empty-database, collapse=TRUE}
# demo database is empty
con %>% DBI::dbListTables()
```

`dbWriteTable()` Creates a remote table from local data.
The last argument is optional.

```{r our-first-table, collapse=TRUE}
# create table by supplying local dataframe
con %>% dbWriteTable("mtcars", mtcars, overwrite = TRUE)
con %>% dbWriteTable("iris", iris, overwrite = TRUE)
# check that tables are properly created
DBI::dbListTables(con)
```

You can also ask the existance of the specific table in the database associated with `con` variable.

```{r , collapse=TRUE}
# boolean question about the existance of indivisual table
con %>% DBI::dbExistsTable("mtcars")
```

`dbListFields()` gets you the list of fields of an indivisual table.

```{r list-fields-mtcars, collapse=TRUE}
# now we have
con %>% DBI::dbListFields("mtcars")
```

`dbReadTable()` allows you to take a quick peep at a table. 

```{r read-table-without-query}
con %>%
    DBI::dbReadTable("mtcars") %>%
    head()
```


You can use `dbSendQuery()` to execute a query written as a string. Note that we don't need a `;` at the end of the query. Since this method does nothing but sending query, you should keep the returned value to perform appropriate post-processing.

```{r first-send-query}
# send query and get response variable
rs <- con %>% DBI::dbSendQuery("
	SELECT *
	FROM mtcars
	WHERE hp <= 100
")
```

`dbFetch()` shows you the result of the query.

```{r first-fetch}
rs %>% DBI::dbFetch()
```

Don't forget to do `dbClearResult()`. 
DBI package keeps many kinds of resources associated with the result of the query unless manually cleared. `dbClearResult()` is responsible for that freeing role.
DBI reference says that 

> This step is mandatory for all objects obtained by calling dbSendQuery() or dbSendStatement().

```{r first-clear-result}
rs %>% dbClearResult()
```

We can write a parameterised query in the following way.
`$1` and `$2` corresponds to the first and second argument in the `list(100, 120)`, respectively. 

```{r first-param-query, collapse=TRUE}
rs <- con %>% dbSendQuery("
	SELECT *
	FROM mtcars
	WHERE hp BETWEEN $1 AND $2
")
rs %>% dbBind(list(100, 120))
rs %>% dbFetch()
rs %>% dbClearResult()
```

You can delete a table by `dbRemoveTable()`. 

```{r remove-tables}
# remove indivisual table by supplying name string
con %>% DBI::dbRemoveTable("mtcars")
```

It is often useful to have a function that deletes multiple tables at once based on a pre-defined rule.
The following code is one of the most simplest, deleting all tables.

```{r, collapse=TRUE}
# create mtcars table again to demonstate sumultaneous-removal
con %>% DBI::dbWriteTable("mtcars", mtcars, overwrite = TRUE)
DBI::dbListTables(con)
# remove multiple tables at once
con %>%
    DBI::dbListTables() %>%
    purrr::map_lgl(~ DBI::dbRemoveTable(con, .x)) %>%
    all()
DBI::dbListTables(con)
```

`dbDisconnect()` terminates the supplied `con` variable, and closes the connection. 

```{r first-dbDisconnect}
con %>% DBI::dbDisconnect()
```